EnVision CF Analyses

### 2025-07-07 Christine Chan/EnVision CF/envision_analyses.qmd
added Amir's analysis to unified Envision analysis

```{r setup}
#| include: false
library(Hmisc)
library(tidyverse)
library(arsenal)
library(pROC)
library(gt)
library(patchwork)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Windows" = "C:/Users/Tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

>>>>

```{r setup}
#| include: false
library(Hmisc)
library(tidyverse)
library(arsenal)
library(pROC)
library(gt)
library(patchwork)
library(lmerTest)
library(emmeans)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Windows" = "C:/Users/Tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

# Hypoglycemia Comparisons

```{r}
# Timepoint and variables lists
timepoints <- c(0, 10, 30, 60, 90, 120, 150, 180)
limited_timepoints <- c(0, 120, 150, 180)
glucoses <- paste0("Glucose_", timepoints)
insulins <- paste0("Insulin_", timepoints)
cpeptides <- paste0("C.Peptide_", timepoints)
epinephrines <- paste0("Epinephrine_", timepoints)
norepinephrines <- paste0("Norepinephrine_", timepoints)
adren_survey <- paste0("adren_score_", limited_timepoints)
neuro_survey <- paste0("neuro_score_", limited_timepoints)
total_survey <- paste0("total_score_", limited_timepoints)
num_symptoms_survey <- paste0("num_symptoms_", limited_timepoints)
hypo_surveys <- c(adren_survey, neuro_survey, total_survey, num_symptoms_survey)
# First visit dataframe
first_visit_df <- final_df %>%
  group_by(study_id) %>%
  filter(row_number() == 1) %>%
  ungroup()
# Long data for mixed models
glmm_df <- final_df %>%
  select(
    study_id, Hypo60, Hypo70,
    all_of(c(
      glucoses, insulins, cpeptides, epinephrines, norepinephrines,
      hypo_surveys
    ))
  ) %>%
  pivot_longer(
    all_of(c(
      glucoses, insulins, cpeptides, epinephrines, norepinephrines,
      hypo_surveys
    )),
    names_to = c(".value", "Timepoint"),
    names_pattern = "(.*)_([0-9]{1,3})"
  )
glmm_df$Timepoint <- factor(glmm_df$Timepoint, levels = timepoints)
```

## Demographics at First Visit

### By Hypoglycemia <70 mg/dL at Any Point

```{r results='asis'}
t1 <- tableby(
  Hypo70 ~
    fe(Diagnosis) + redcap_data_access_group +
    kwt(age_visit, "Nmiss", "medianq1q3", "range") +
    kwt(height, "Nmiss", "medianq1q3", "range") +
    kwt(weight, "Nmiss", "medianq1q3", "range") + bmi_perc + sex +
    fe(origin_race) + fe(ethnicity) + fe(pancreatic_status) + fe(CFTR) +
    corrector_yes_no + fev1 + fvc + a1c_result,
  data = first_visit_df, control = tableby.control(simulate.p.value = T)
)
summary(t1, pfootnote = T)
```

### By Hypoglycemia <60 mg/dL at Any Point

```{r results='asis'}
t2 <- tableby(
  Hypo60 ~
    fe(Diagnosis) + redcap_data_access_group +
    kwt(age_visit, "Nmiss", "medianq1q3", "range") +
    kwt(height, "Nmiss", "medianq1q3", "range") +
    kwt(weight, "Nmiss", "medianq1q3", "range") + bmi_perc + sex +
    fe(origin_race) + fe(ethnicity) + fe(pancreatic_status) + fe(CFTR) +
    corrector_yes_no + fev1 + fvc + a1c_result,
  data = first_visit_df, control = tableby.control(simulate.p.value = T)
)
summary(t2, pfootnote = T)
```

## Glucose

Glucose was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable. The results from this model are presented as estimated marginal means (also known as least-squares means) for those with hypoglycemia at some point and those with no hypoglycemia.

### Hypoglycemia <70 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(Glucose ~ Timepoint * Hypo70 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean Glucose") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean glucose at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour glucose was `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour glucose was `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`).

### Hypoglycemia <60 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(Glucose ~ Timepoint * Hypo60 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean Glucose") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean glucose at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour glucose was `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour glucose was `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`).

## Insulin

Insulin was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable. The results from this model are presented as estimated marginal means (also known as least-squares means) for those with hypoglycemia at some point and those with no hypoglycemia.

### Hypoglycemia <70 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(Insulin ~ Timepoint * Hypo70 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean Insulin") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean insulin at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour insulin was `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour insulin was `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`).

### Hypoglycemia <60 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(Insulin ~ Timepoint * Hypo60 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean Insulin") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean insulin at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour insulin was `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour insulin was `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`).

## C-peptide

C-peptide was modeled using a linear mixed effect model with random intercept for participant and time treated as a categorical variable. The results from this model are presented as estimated marginal means (also known as least-squares means) for those with hypoglycemia at some point and those with no hypoglycemia.

### Hypoglycemia <70 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(C.Peptide ~ Timepoint * Hypo70 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo70"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo70", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo70,
  group = Hypo70
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean C-peptide") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean C-peptide at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour C-peptide was `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="No Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour C-peptide was `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo70=="Hypoglycemia < 70" & means$Timepoint=="120","CI Upper Limit"],2)`).

### Hypoglycemia <60 mg/dL at Any Point

```{r}
# Fit
mod <- lmer(C.Peptide ~ Timepoint * Hypo60 + (1 | study_id),
  data = glmm_df
)
# EM means
means <- data.frame(emmeans(mod, c("Timepoint", "Hypo60"),
  type = "response",
  pbkrtest.limit = 1e6
))
colnames(means) <- c(
  "Timepoint", "Hypo60", "Mean", "SE", "df", "CI Lower Limit", "CI Upper Limit"
)
# Plot
ggplot(means, aes(
  x = as.numeric(as.character(Timepoint)), y = Mean, colour = Hypo60,
  group = Hypo60
)) +
  geom_point() +
  geom_errorbar(aes(ymin = `CI Lower Limit`, ymax = `CI Upper Limit`),
    width = 3
  ) +
  geom_line() +
  scale_x_continuous(breaks = timepoints) +
  scale_color_manual(values = c("#66C2A5", "#FC8D62")) +
  ylab("EM Mean C-peptide") +
  xlab("Timepoint") +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank())
```

The plot above shows the estimated mean at each time point. Error bars indicate the 95% confidence interval.

```{r}
means %>%
  gt() %>%
  fmt_number(decimals = 2)
```

The table above shows the estimated mean C-peptide at each time point for the two groups. For example, for those with no hypoglycemia the mean two hour C-peptide was `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="No Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`) and for those with hypoglycemia at some point the mean two hour C-peptide was `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","Mean"],2)` (95% CI: `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Lower Limit"],2)` - `r round(means[means$Hypo60=="Hypoglycemia < 60" & means$Timepoint=="120","CI Upper Limit"],2)`).
